Active Code Completion

### 2012-03-18 ecoop12/rebuttal.txt

We thank the reviewers for their thorough critiques and thoughtful questions. We would like to particularly thank Reviewer 3 for soliciting an additional review from a researcher in the targeted user community.

Reviewer 1 expressed concern about the originality of Section 2. We agree that Section 2 does not introduce any individually novel design or adoption criteria. The purpose of this section was to bring together the issues important to language designers in this domain, which has never been done before. We do this to support our position that language designs should be evaluated holistically, against clear design and adoption criteria, consistent with the practices of other complex design disciplines. We will expand on this justification at the beginning of Section 2 in the final manuscript.

We acknowledge the concern raised by reviewers that Section 3 does not clearly state which aspects of the language design are individually novel. We believe that several of the language mechanisms we describe are in fact novel:

* The structural type inference scheme that cl.oquence uses has, to our knowledge, not been previously described. As described in the paper, within-function type inference is combined with type propagation between functions. Aspects of the within-function inference are controlled by the extension mechanism.
* To our knowledge, no language for GPU computing has provided kernel inference and no language targeting OpenCL has provided extension inference.
* The approach described for providing higher-order functions that are not first-class by inlining, rather than by the use of function pointers, has to our knowledge not been explored in existing languages (compilers sometimes do this, but languages do not guarantee this behavior.)
* Active libraries, although proposed previously, have not been adequately explored in practical languages (previous approaches relied on C++ template expansion). We expand on the concept of active libraries significantly, showing how a rich metalanguage (Python) enables a number of interesting language features (Sections 3.5-3.8).
* The language extension mechanism we describe is, to our knowledge, entirely novel, and is a major contribution of this paper.

We will take care to emphasize the extent to which each feature we describe is novel in the final manuscript.

Several reviewers note that we did not provide benchmarks to justify our claim that the OpenCL backend is performance-neutral. In fact, this claim is not an empirical one. We justify it by noting that all of the mechanisms that we describe are compile-time mechanisms -- cl.oquence does not introduce any additional run-time machinery or code transformations that may be the source of performance degradation. However, we would be happy to provide our benchmark data that demonstrates this 100% parity in performance to further emphasize this point in the final manuscript.

Several reviewers asked for more specific references to design criteria, in particular in section 3.2. We acknowledge this lack of specificity. Although we do reference section headings whenever relevant, we will be more specific about the way in which the described feature satisfies the referenced criteria in the final manuscript.

### 2012-03-18 vlhcc11/survey-part.tex
Initial paper move.

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt, conference, compsocconf]{IEEEtran}
% Add the compsocconf option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

\usepackage{rotating}

@@

@@

\usepackage{alltt}
\usepackage{hyperref}
\hypersetup{pdfborder={0 0 0 0}}

@@

% author names and affiliations
% use a multiple column layout for up to two different
% affiliations

\author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
\IEEEauthorblockA{Carnegie Mellon University, 
Pittsburgh, PA, USA\\
\small{\texttt{\{comar,youngseok,tlatoza,bam\}$@$cs.cmu.edu}}\\
\small{\texttt{\url{http://www.cs.cmu.edu/~NatProg/graphite.html}}}}
}

@@

% make the title area
\maketitle

\begin{abstract}
Code completion menus have replaced standalone API browsers for most developers because they are more tightly  integrated into the development workflow. Refinements to the code completion menu that incorporate additional sources of information have similarly been shown to be valuable, even relative to standalone counterparts offering similar functionality. In this paper, we describe {\em active code completion}, a technique that enables the integration of a much broader array of interactive and highly-specialized code generation interfaces, called {\em palettes}, directly into the editor environment. Using several empirical methods, we examine the contexts in which such a system could be useful, describe the design constraints governing the system architecture as well as particular code completion interfaces, and design one such system, named Graphite, for the Eclipse Java development environment. Using Graphite, we implement a palette for writing regular expressions as our primary  example and conduct a controlled user study that provides further evidence in support of the claim that integrating specialized tools directly into the editor environment is both feasible, particularly if aided by an  active code completion system like Graphite, and valuable to professional developers.
\end{abstract}

\begin{IEEEkeywords}
code completion, development environments

\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
Software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}. Most editors provide code completion in the form of a floating menu containing  contextually-relevant variables, fields, methods, types and other code snippets. By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool or API browser.
% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.

Several refinements and additions to the code completion menu have previously been suggested in the literature. These have focused on leveraging additional sources of information, such as databases of usage history \cite{robbes_how_2008}, examples extracted from code repositories \cite{bruch_learning_2009} and crowdsourced information \cite{mooty_calcite:_2010}, to increase the relevance and sophistication of the featured menu items. As with the standard form of code completion, many of these sources of data can also be utilized via external tools (e.g. Calcite \cite{mooty_calcite:_2010} uses information that could already be accessed using the Jadeite \cite{conf/vl/StylosFYM09} tool). Empirical evidence presented in these studies, however, indicates that directly integrating these kinds of tools directly into the editor environment can be more effective.

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also limited in that they are difficult to modularly extend. A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

In this paper, we discuss active code completion in the context of object construction. For example, consider the following Java code fragment:

\begin{alltt}
  public Color getDefaultColor() \{
      return \textvisiblespace
\end{alltt}

\begin{figure*}\label{color}
\begin{center}
\includegraphics[scale=0.8]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The source code generated by this palette.}
\end{figure*}

If the developer invokes the code completion command at the indicated cursor position, the editor  looks  for a {\it palette definition} associated with the {\it type} of the expression being entered, which in this case is  \verb|Color|. If an associated palette is found, a menu item briefly describing this palette is added to the standard code completion menu. When selected, the corresponding palette is shown, replacing the standard code completion menu. Figure 1a gives an example of a simple palette that may be associated with the \verb|Color| class.

The developer can interact with such palettes to provide parameters and other information related to her intent, and receive immediate feedback about the effect these choices will have on the behavior of the object being constructed. When this interaction is complete, the palette is responsible for generating appropriate source code for insertion at the cursor. Figure 1b shows the code that is inserted after the user presses the \textsc{enter} key.

In accordance with rational design practices, we sought to address the following questions before designing and implementing our active code completion system:

\begin{itemize}
\item What {\it specific} use cases exist for this form of active code completion in a professional development setting? 
\item What {\it general} criteria are common to types that would and would not benefit from an associated palette?
\item What are some relevant usability and design criteria for palettes designed to address such use cases?
\item What capabilities must the underlying active code completion system provide to enable these use cases and user interface designs?
\end{itemize}

To help us answer these questions, we conducted a large online survey of professional developers (Section 2). Their responses, along with information gathered from informal interviews and code corpus analyses, revealed a number of interesting use cases (Section 3) and non-trivial functional requirements for palette interfaces as well as the underlying active code completion architecture (Section 4). Notably, many of these findings can be applied to development tools for professional developers more generally, not simply those integrated into the editor in the way we have shown.

Next, we describe Graphite, an Eclipse plug-in that provides active code completion for users of the Java programming language (Section 5). We describe several interesting design choices that were made to satisfy many of the requirements discovered in our preliminary investigations and briefly examine the trade-offs that were necessary due to Java's purely textual program representations and Eclipse's limited extension APIs.

Finally, we conduct a controlled lab study with a more complex palette, implemented using Graphite, that assists  developers as they write regular expressions (Section 6). The data and observations gathered from this study provide  specific evidence in support of the broader claim that highly-specialized tools that are integrated directly with the editing environment are particularly useful, and that an active code completion system like Graphite makes developing, deploying and discovering such tools fundamentally simpler.

\section{Survey Results}
To validate our general conceptualization of active code completion, develop concrete design criteria and create a concrete list of use cases, we began by conducting a large survey of professional software developers.

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
\subsection{Survey Methods}
We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com\footnote{http://www.reddit.com/r/programming}. An additional 22 participants were recruited from the local computer science graduate student body.

Recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.
Participants were told that the survey would take approximately 20 minutes to complete, and no reward was offered. Of the 696 people who started the survey, 473 participants (68\%) completed it. We examine the responses from completed surveys only in the analyses below.

@@

We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''. 61.1\% of the participants indicated that they were an expert in at least one language, and an additional 35.7\% were ``very familiar'' with at least one language. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.

We also asked participants to select which integrated development environments (IDEs) and code editors that they were familiar with. The Eclipse IDE was familiar to 87.1\% of participants. This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%. Participants could also enter ``other'' choices and a number of editors and IDEs were entered, including Xcode, Textmate and Notepad++.

\subsection{Palette Mockups}

%\begin{figure}
%  \includegraphics[width=\linewidth]{mockup-palette-small.png}
%  \caption{An example mockup palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
%  \label{regex}
%\end{figure}
%
Next, we presented participants with a series of mockup palettes for a Color class (more complex than the one we ultimately implemented in Figure 1a), a regular expression class, and a SQL query class. Participants were also shown mockup screenshots demonstrating how a user could invoke the palette, and a mockup showing the code that would be inserted once a selection had been made. Before presenting each mockup, we gathered information about  the strategies that they would normally use to instantiate the class.

For the Color class, the majority of participants indicated that they would look in the code completion menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant if asked to instantiate an object corresponding to the color ``navy'' (which is not, in fact, a standard color in Java.) Another 14.0\% indicated that they would use an external tool (such as an image editor) to determine the RGB values corresponding to the color.

Before asking participants about regular expressions and SQL queries, we asked participants to rate their familiarity with these concepts. Few participants (4\%) indicated that they were unfamiliar with regular expressions and no participants were unfamiliar with creating SQL queries, providing further evidence that our participants were not novice developers. Fig. \ref{strategies} summarizes the strategies that participants generally preferred for instantiating regular expressions and SQL queries. Using an external tool was a common strategy in both cases, particularly for SQL queries, but several other strategies were also represented.

\begin{figure}
\begin{tabular}{lccc}
 & Regular Expressions &  & SQL\\
 \hline
Separate test script & 29.6\% &  & 15.4\%\\
Guess and check & 14.0\% &  & 16.1\%\\
External tool & \textbf{37.9\%} &  & \textbf{58.6\%}\\
Search for examples & 12.3\% & & 5.1\% \\
Other & 6.2\% & & 4.9\% \\
\hline
\end{tabular}
\caption{Distribution of responses to survey questions asking about typical strategies for writing regular expressions and SQL queries.}
\label{strategies}
\end{figure}

@@

Finally, after showing the series of mockup screenshots, we asked participants to rate how useful the integrated palette would be to them if they needed to instantiate the corresponding class. The responses to this question for each palette are summarized in Fig. \ref{frequencies}. In each case, more than half of the participants indicated that they would use active code completion at least some of the time. The regular expression palette was considered particularly useful while the color and SQL palettes showed a more reserved pattern of responses.

In addition to asking for a simple rating for each palette, we also solicited open-ended comments. A large number of participants volunteered comments: 193 for the color palette, 129 for the regular expression palette and 142 for the SQL palette. These responses were highly valuable when developing the design criteria below and helped to explain the patterns observed in Fig. \ref{frequencies}.

\section{Design Criteria}
Using the information gathered from the survey as well as informal discussions with developers and researchers over the course of this study, we sought to develop design criteria constraining both the overall system design as well as the design of individual palettes. In the section headings below, the number of survey responses, summed over the three palette mockups, that contained the listed concern, as judged by the authors of this paper, are listed in parenthesis. Many of these concerns apply generally to various kinds of editor-integrated tools, not just those implemented using the active code completion mechanism that we focus on, so we hope these design criteria will be valuable to the developer tool design community more broadly.

\subsection{Maintaining Separation of Concerns (183)}
The most common issue developers had, particularly with the Color palette, was that it seemed to ignore the issue of separation of concerns by allowing developers to insert color constants directly into the program logic. Many developers noted that this is considered bad practice, or should be limited to the prototyping phase of a project. This concern was also expressed in responses to the SQL palette, which required inserting parameters to connect to a database so that the query could be tested. The code that was inserted included initialization steps needed to connect to the specific database that was entered, and several participants noted that much of this information should appear in an external resource file rather than inline with the program logic itself. Few participants made similar comments about the regular expression palette, however, indicating that regular expressions are considered a part of the program logic rather than data by most developers.

This feedback suggests that tool and palette designers may wish to explicitly aprise users of relevant best practices and acknowledge that palettes that generate constant data may be most useful in the prototyping phase. It can be noted that when transitioning from a prototype to production-quality software the code generated by a palette or tool may be used as template that can be refactored to follow stricter policies as needed. We showed users the Color and SQL palettes in the context of a standard programming editor, but these comments indicate that resource file and  stylesheet editors may benefit from active code completion support as well.

\subsection{Integration with Testing Frameworks (35)}
The regular expression palette shown to the participants allowed users to immediately test a pattern against provided strings. These test strings and the results of performing the match were inserted as comments below the generated source code. A number of participants requested that explicit unit tests be generated instead. Although the palette interface conveyed the results of these tests immediately, participants remained adamant about explicit unit tests, likely due to concerns that future modifications to the generated code without reinvoking the palette might introduce bugs, or due to the desire to conform to standard testing practices. To support the generation of unit tests, however, the active code completion architecture must explicitly allow for code generation at locations other than at the cursor and handle the fact that statements generated after a code completion command are not necessarily encapsulated well enough to be immediately testable in isolation.

@@

\subsection{Support for Reinvocation (19)}
Several participants asked for the ability to reinvoke a palette from previously generated source code. In order to support this feature, the architecture must provide the palette with enough information to reconstruct its state. To complicate matters, however, users may wish to modify the generated code between invocations of the palette and have these modifications reflected in the palette's state upon reinvocation (e.g. modify the RGB values in the case of the \verb|Color| class). Moreover, there may be important aspects of the palette's state that are not directly available in the generated code, such as parameters controlling the palette's user interface. Indeed, associating tool-related metadata with code is known to be cumbersome in purely textual languages, since all metadata must be directly visible within comments or annotations.

\subsection{Support for Palette Settings and History (41)}

A related feature important to many participants is support for maintaining settings and usage history across invocations of the same palette at different code locations. For example, 20 participants requested that the Color palette include a list of recent or favorite colors, and 12 participants specifically inquired about whether the database connection information was maintained between invocations of the SQL palette.

% The followings are the comments that are related to this category.

@@

\subsection{Support for Nested Expressions (13)}

In all of the examples that we gave, the parameters entered into the palette interface corresponded to simple, constant expressions, rather than complex expressions referring to variables from the surrounding context. A number of participants noticed this limitation. For example, several participants asked SQL query strings, as these are typically constructed using user-generated data in practice. Although a simple expression entry box may suffice in simple scenarios, architectural support is needed for palettes that need to inspect the code context (e.g. to verify well-formedness) or if code highlighting, code completion and other advanced editing features are needed within the palette itself.

\subsection{Keyboard Navigability (12)}

Although our presentation of active code completion did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse. These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were familiar with editors like Vim that place a strong emphasis on keyboard shortcuts.

\subsection{Minimal Impact on Editor Responsiveness}

A common theme in our discussions with developers (as well as in comments left on our recruitment thread) was that integrated development environments like Eclipse were already too slow, and that an extension such as the one we were proposing would only be acceptable if it did not affect performance and responsiveness any further.

\subsection{IDE and Language Portability}

The mockups we showed users were based on the Java language and the Eclipse development environment. As we showed, a number of participants preferred other languages or editors. Many of these participants made comments asking that the features we describe not be restricted to one specific editor (e.g. NetBeans users) and some also called for support for other programming languages (e.g. Ruby and C\#). Indeed, the palettes we demonstrated could be used with only slight modifications in a variety of programming languages, given suitable architectural support for porting palettes between editing environments.

\subsection{Varying User Needs}

Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically saying that they would be satisfied with a simpler interface with colors listed next to their names. Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen. Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively. User requests included syntax highlighting, a range of additional helpful tools for generating,  testing and sharing regular expressions, and the ability to browse databases more directly. Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice. To support incremental improvements to palettes based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.

\section{Use Cases}
At the end of the survey, we solicited suggestions from the participants of other classes that could benefit from an associated palette. 119 participants made one or more suggestion and we classified the suggestions into several categories. We also mention other suggestions made by researchers and developers via other channels in some cases, without including them in the provided counts.

\subsection{Graphical Elements (27)}
The most popular suggestions were graphical elements, influenced perhaps by our demonstration of the Color palette. Some participants suggested palettes for classes representing primitive graphical objects, such as brush and font selectors or polygon editors, while other participants were focused on user interface elements, such as buttons, check boxes and frame layouts. A few also suggested palettes for manipulating 3D primitives, such as transformation matrices, in a more direct and intuitive manner. A practitioner also suggested that because setting up a plot or graph is often significantly simpler using a direct manipulation interface, it would be a natural candidate for a palette as well.

\subsection{Query Languages (17)}
The second most popular category of suggestions consisted of various interfaces for query languages, also likely due to the examples we provided to participants. In addition to variants of the SQL and regular expression palettes, developers also wanted to work with other types of queries such as XPath or XQuery for XML.

\subsection{Simplified or Domain-Specific Syntax (16)}
Another interesting class of suggestions were cases where a more natural syntax than the syntax provided by Java is desirable. One suggestion was a palette that automatically escaped strings containing quotation marks or escape sequences. A related category of suggestions consisted of palettes that offered a more natural interface for generating strings containing code in other languages such as HTML (e.g. offering syntax highlighting, escaping,  tag matching and other features.) Domain-specific syntax for complex mathematical expressions and chemical formula were also mentioned in discussions with practitioners.

An interesting suggestion that we investigated further involved Java's collection classes, such as \verb|ArrayList| and \verb|HashMap|. A participant suggested that these classes could be associated with a palette that offered a simplified literal syntax for initialization, pointing toward other languages that do offer such a literal syntax (e.g. JavaScript.) Without such syntax, these classes must be tediously initialized using a separate method call for each element. To determine whether this usage pattern is common, we conducted a corpus analysis using 10 randomly selected projects from the Qualitas Corpus \cite{QualitasCorpus:APSEC:2010} containing over 1M lines of code. We began by searching for places in these projects where Java collection classes were being instantiated, then looked to see whether this instantiation code was immediately followed by method calls that inserted items into the collection, indicating a case where a literal may have been used if available. Figure 4 summarizes the results of this analysis, providing evidence in support of the claim that a palette that simplifies this process could be useful for general-purpose programming.

\subsection{Unclear Parameter Implications (11)}
Another category of use cases contains classes where it can be difficult to predict what the run-time behavior of a particular parameter choice may be. Examples given included audio filters (e.g. pitch manipulation) and animation descriptors (e.g. speed or shape parameters). By giving immediate visual or auditory feedback using a preview panel, these parameters can be tweaked without requiring the execution of the full application.

%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item String
%		\item Collections (e.g., dictionary)
%		\item Vector, Matrix
%		\item Embedded languages
%		\item URLs, Paths
%	\end{itemize}
%	
%	\item Expected features
%	
%	\begin{itemize}
%		\item Proper escaping
%		\item Syntax highlighting
%		\item Auto-completion
%	\end{itemize}
%\end{itemize}

\begin{figure}\label{collections}
\begin{center}
\begin{tabular}{l|c|c|r}
Collection Class& Total & Literal & Percentage \\
\hline
\texttt{ArrayList} & 464	&	44	&	9.5\% \\
\texttt{HashMap}	& 56 & 19 & 33.9\% \\
\texttt{HashSet} & 122 & 62 & 50.8\% \\
\texttt{Hashtable} & 86 & 10 & 11.6\% \\
\texttt{Vector} & 729 & 31 & 4.2\% \\
\hline
\textbf{Total} & 1457 & 166 & 11.4\% \\
\hline
\end{tabular}
\caption{Usage patterns for common Java collection classes in the \texttt{java.util} package in our code corpus. Uses that fit a pattern that can be captured by a literal make up a significant portion of all uses. Not all possible usage scenarios of this type were captured by our analysis, so these numbers are lower bounds.}
\end{center}
\end{figure}

@@

@@

@@

\subsection{Integrating with Documentation and Examples (7)}
Some participants suggested integrating tutorials or lists of relevant examples directly using a palette, so that these can be discovered more easily by new users and inserted directly into code, without requiring switching to a web browser and executing a search.

\subsection{Complex Instantiation and Cleanup Procedures (5)}
A related category contains classes that require complex instantiation and cleanup procedures. For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class. This class can be difficult to use because it requires try/catch block, and one must remember to close the file after reading it. By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects and get an outline containing the full life-cycle of the file. Similarly, palettes may help to alleviate the factory pattern usability problem \cite{ellis_factory_2007}. As long as the developers remember which class to use, they will not need to remember how to instantiate that class. We explore this further in our user study in Section VI.

@@

%\section{Tool}
%Described earlier.
%\section{Developer Survey}
%
%\subsection{Methods}

>>>>

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

\usepackage{enumerate}
\usepackage{rotating}
\usepackage{url}

@@

@@

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Cyrus Omar, YoungSeok Yoon, Thomas D. LaToza, Brad A. Myers}
\IEEEauthorblockA{School of Computer Science\\
Carnegie Mellon University\\
Pittsburgh, PA, USA\\
\{comar,youngseok,tlatoza,bam\}@cs.cmu.edu}
\texttt{\url{https://github.com/cyrus-/graphite}}
}

@@

% make the title area
% \maketitle

% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
% \IEEEpeerreviewmaketitle

\section{Developer Survey}

Prior to implementation, we conducted a large online survey of professional software developers in order to extract criteria that govern whether our system would be useful to them. More specifically, we sought to address the following questions:

\begin{itemize}
\item What are specific use cases for active code completion in a professional setting? 
\item What are some functional criteria that predict whether a class would benefit from active code completion?
\item What are some usability criteria that should govern palette interface designs?
\item Which capabilities must a active code completion architecture support to enable these use cases and designs?
\end{itemize}

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mock-up screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.

We posted a link to our survey on a popular  collaborative filtering website \cite{mehta_robust_2007} called {\tt /r/programming} on reddit.com\footnote{\texttt{http://www.reddit.com/r/programming}}. At the time of our study, the website listed approximately 340,000, though members were not required to registered to read the content,  and contained content that could be considered relevant to our intended audience. An additional 22 participants were recruited using a mass email to local computer science graduate students. The recruitment materials in both cases stated that we were seeking developers ``familiar with an object-oriented programming language like Java, C\# or Visual Basic and an integrated development environment like Eclipse or Visual Studio''.

Participants were asked to take an online survey that would take approximately 20 minutes to complete. Of the 696 people who started the survey by answering the first question, 473 participants completed it. Of these, 16 participants were computer science graduate students at Carnegie Mellon and 457 were from \texttt{/r/programming}.

Their responses revealed a number of interesting use cases and non-trivial design constraints for active code completion systems. Many of these findings can also be generalized to constrain the design of visual programming languages, particularly those that aim be useful to professional developers, rather than novice programmers (e.g., Barista \cite{ko_barista:_2006}). We summarize the survey questions and the results in the following subsections.
% In fact, active code completion can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional representations of program logic. This may help address some of the usability challenges previously associated with visual languages (cf. [?]).

@@

We first asked their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''. 61.1\% of the respondents expressed themselves as experts on at least one language, and 35.7\% were very familiar with at least one language. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.

We also asked participants to select the integrated development environments (IDEs) and code editors that they were familiar with. The Eclipse IDE was familiar to 87.1\% of participants. This was followed by Visual Studio at 66.0\%, Vi/Vim at 53.7\%, Netbeans at 37.7\%, Emacs at 24.8\% and IntelliJ IDEA at 16.4\%. Participants could also enter ``other'' choices. A number of editors and IDEs were entered, including Xcode, Textmate and Notepad++. None of these were dominant.

\subsection{Mock-ups}

\begin{figure}
  \includegraphics[width=\linewidth]{mockup-palette-small.png}
  \caption{An example mock-up palette for regular expressions shown to the survey participants. In the survey, they were shown the surrounding Eclipse IDE together.}
  \label{regex}
\end{figure}

Next, we presented participants with a series of mock-up palettes for a Color class, a regular expression class (Fig. \ref{regex}), and a SQL query class. Participants were also shown a mock-up demonstrating how a user could invoke the palette, and a mock-up showing the code that would be inserted once a selection had been made.

Before presenting each mock-up, we gathered information about the user's familiarity with the class and the strategies that they would likely use to instantiate the class under normal circumstances. For the Color class, the majority of participants indicated that they would look in the code completions menu (58.4\%) or in the class documentation (19.0\%) for a predefined constant. Another 14.0\% indicated that they would use an external tool to determine the RGB values. Few participants were unfamiliar with regular expressions and no participants were unfamiliar with SQL queries which provides further evidence that our participants were not novice developers. Fig. \ref{strategies} summarizes their usual strategies for instantiating regular expressions and SQL queries.

\begin{figure}
\begin{tabular}{lccc}
 & Regular Expressions &  & SQL\\
 \hline
Separate test script & 29.6\% &  & 15.4\%\\
Guess and check & 14.0\% &  & 16.1\%\\
External tool & \textbf{37.9\%} &  & \textbf{58.6\%}\\
Search for examples & 12.3\% & & 5.1\% \\
Other & 6.2\% & & 4.9\% \\
\hline
\end{tabular}
\caption{Typical strategies for regular expressions and SQL queries.}
\label{strategies}
\end{figure}

@@

Participants were positive about our mock-ups. We asked the users to rate the usefulness of each mock-up palette, and their responses are summarized in Fig. \ref{frequencies}. For each palette, more than half of the respondents expressed that they would use it at least some of the time. Especially, people liked the regular expression palette a lot.

\section{Design Implications}

We analyzed the open-ended responses to extract design principles and constraints. We got 193 responses for the color palette, 129 for the regular expression palette, and 142 for the SQL palette. Also, we got total 119 suggestions about what other types of classes could benefit from similar types of palettes. The extracted principles and constraints are summarized in the following subsections. The parenthesized number following each design constraint indicates how many people commented on that issue.

\subsection{System Design Constraints}

System design constraints are the criteria that should be taken into account in the design of the overall system architecture.

\subsubsection{Handling Separation of Concerns (183?)}

Participants were very wary of including constant data into the source code directly. Many responses expressed general sentiment or specific preference for project-wide color theme or external resource file. Sometimes they explicitly suggested to add capability to handle the external resource files directly from the palette. Interestingly, most people seemed to consider regular expressions as part of the program logic. There were few people who complained about including the regular expression pattern string in the source code as opposed to the other two classes.

@@

\subsubsection{Reversibility (19?)}

Some participants expressed that it should be able to bring the palettes back when they want to modify the parameters after they used the palettes.

\subsubsection{Palette Settings and State (41?)}

Many participants wanted to be able to configure the palettes or have the palettes maintain state even after the palettes are hidden. % The followings are the comments that are related to this category.

@@

\subsubsection{Interaction with Code Context (13?)}

Some people wanted to interact with the code context. For example, people expressed that it would be great if the SQL query palette was capable of dynamically creating SQL query string by combining multiple string variables, or by assigning variables for the parameters.

\subsubsection{Performance (?)}

Developers were concerned about the performance of these palettes. Some of the participants said that they would use these palettes only if it does not slow down the IDE. Also, this was the most popular comment on the reddit thread.

\subsubsection{IDE Independence (?)}

Several participants expressed a desire for IDE or even language independence for these palettes.

\subsection{General Palette Design Considerations}

General palette design considerations are about how to make more usable palettes.

\subsubsection{Simplicity vs. Capability}

Even though the participants were shown the same mockup screenshots, their reactions were split. Our color palette was considered too complex by many participants (26), and many others (26) would be satisfied with seeing colors next to a list of color names. The regular expressoin palette and SQL palette were considered too simple (12, and 15 participants respectively). They wanted syntax highlighting, match highlighting, and browsing capability of SQL databases.

\subsubsection{Keyboard Navigability (12)}

12 of the participants expressed strong antipathy toward the mouse. This was mostly directed at the color palette.

\subsection{Suggestions from the Participants}
At the end of the survey, we solicited suggestions from the participants about what types of classes could benefit from the active code completion palettes. 119 participants suggested various types of classes and we classified the suggestions into several categories.

\subsubsection{Alternative/Tricky/Literal Syntax (16)}
These are the classes of which syntax is tricky, or complex. Even for a basic string class, people wanted to input multiline string or unicode string in more convenient way with automatic escaping. We address this issue in more detail in Section \#.
%
%\begin{itemize}
%	\item Example classes
%	
%	\begin{itemize}
%		\item String
%		\item Collections (e.g., dictionary)
%		\item Vector, Matrix
%		\item Embedded languages
%		\item URLs, Paths
%	\end{itemize}
%	
%	\item Expected features
%	
%	\begin{itemize}
%		\item Proper escaping
%		\item Syntax highlighting
%		\item Auto-completion
%	\end{itemize}
%\end{itemize}

\subsubsection{Unclear Parameter Implications (11)}
The classes in this category has some parameters in it, and it is not easy to predict the results of the parameter modifications. Instead of running the whole application to see the result, the palette can serve as a preview panel so that the developers may check the results directly. Also, the palette can be a control panel which facilitates modifying the parameters as in Juxtapose [?].

@@

\subsubsection{Query Languages (17)}
Similar to the regular expression and SQL palette we presented, the developers also wanted to check if the other types of query strings (e.g., XPath / XQuery) are correct or not by checking the results directly on the palette.

@@

\subsubsection{Graphical Elements (27)}
The other popular category was graphical elements. We've already shown the color example, and the participants suggested many other graphical properties such as font, brush, shape and others. Also, people wanted to check or directly manipulate the GUI frames and layouts.

@@

\subsection{Complex Instantiation Procedures (5)}
Some of the participants pointed out that these palettes might be useful to create an object which requires complex instantiation code. For example, in order to read a text file in Java, the developer might want to use \texttt{BufferedReader} class. It is error prone to use this class because it requires try/catch block, and also one must close the file after reading it. By using a palette to choose a file or choose a variable which contains the file path, the developers could easily instantiate these objects. In the same way, it can alleviate the factory pattern usability problem \cite{ellis_factory_2007}. As long as the developers remember which class to use, they will not need to remember how to instantiate that class.

@@

\subsection{Integrating with Documentation, Tutorial (7)}
Some participants suggested integrating the documentation or the tutorials into the interactive palettes. If the developers are not familiar of the specific API class, the palette could tell them how to use it.
